---
title: "Tidying"
output: pdf_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
collegeAdmission <- read_csv("~/Stats 272 F22/Project/Group 7A: Swagat, Zoe, Maya/CollegeAdmissions.csv")
```

```{r}
#Getting the right vars and renaming
cleanedDf <- collegeAdmission[c(1, 33, 38, 41, 42, 43, 44, 49, 97, 64:72, 74, 100, 103)] # 8, 10, 12)] 
cleanedDf <- cleanedDf[c(1,8,9, 2:7, 10:21)]


cleanedDf <- rename(cleanedDf, 
       undergradEnrollment = `Undergraduate enrollment`,
       freshmenFinAidPercent = `Percent of freshmen receiving any financial aid` ,
        tuitionFees2013_14 = `Tuition and fees, 2013-14`, 
       geographicRegion = `Geographic region`, 
       control = `Control of institution`, 
       HBCU = `Historically Black College or University`, 
       degreeUrbanization = `Degree of urbanization (Urban-centric locale)`, 
       perc_AmericanIndianAlaskaNative = `Percent of undergraduate enrollment that are American Indian or Alaska Native`, 
       perc_Asian = `Percent of undergraduate enrollment that are Asian`, 
       perc_BlackAfricanAmerican = `Percent of undergraduate enrollment that are Black or African American`, 
       perc_HispanicLatino = `Percent of undergraduate enrollment that are Hispanic/Latino`, 
       perc_NativeHawaiian_OtherPacificIslander = `Percent of undergraduate enrollment that are Native Hawaiian or Other Pacific Islander` , 
       perc_White = `Percent of undergraduate enrollment that are White` , 
       perc_MutiRacial = `Percent of undergraduate enrollment that are two or more races` , 
       perc_RaceUnknown = `Percent of undergraduate enrollment that are Race/ethnicity unknown` , 
       perc_NonresidentAlien = `Percent of undergraduate enrollment that are Nonresident Alien`, 
       perc_Women = `Percent of undergraduate enrollment that are women`,
       perc_FreshmenPellGrants = `Percent of freshmen receiving Pell grants`, 
       perc_FreshmenInstGrantAid = `Percent of freshmen receiving institutional grant aid`, 
     #  SATMath75 = `SAT Math 75th percentile score`, 
      # SATReading75 = `SAT Critical Reading 75th percentile score`, 
       #SATWriting75 = `SAT Writing 75th percentile score`
       )

```

```{r}
#mutating the variables we need
cleanedDf <-cleanedDf %>%
  mutate(
    numPOC = 
      (perc_AmericanIndianAlaskaNative/100 + 
         perc_Asian/100 + 
         perc_BlackAfricanAmerican/100 + 
         perc_HispanicLatino/100 + 
         perc_NativeHawaiian_OtherPacificIslander/100 ) * undergradEnrollment, 
    perc_POC = 
      numPOC/undergradEnrollment*100,
    
    numWhite = perc_White/100 * undergradEnrollment,
  #  SATReading75_10 = SATReading75/10, 
   # SATWriting75_10 = SATWriting75/10,
  #  SATMath75_10 = SATMath75/10
  )

cleanedDf <- cleanedDf %>%
  mutate(tuition1000Dollar = tuitionFees2013_14/1000,#Transform tuition into 1000's of dollars
         controlPublic = ifelse(control == 'Public', 1, 0), 
         HBCUYes = ifelse(HBCU == 'Yes', 1, 0)
         ) 
```

There are a lot of levels for degree of urbanization, so condensing them might be easier for interpretation and regression. 
```{r}
#condense the categories for degree of urbanization
cleanedDf <- cleanedDf %>%
  mutate(
    degreeUrbanization = ifelse(degreeUrbanization == 'City: Midsize'| degreeUrbanization == 'City: Small'| degreeUrbanization == 'City: Large', 'City', degreeUrbanization), 
    
    degreeUrbanization = ifelse(degreeUrbanization == 'Town: Fringe'|
                                  degreeUrbanization == 'Town: Distant'|
                                degreeUrbanization == 'Town: Remote', 'Town', degreeUrbanization), 
    
     degreeUrbanization = ifelse(degreeUrbanization == 'Rural: Remote'|
                                  degreeUrbanization == 'Rural: Fringe'|
                                degreeUrbanization == 'Rural: Distant', 'Rural', degreeUrbanization), 
    
    
    degreeUrbanization = ifelse(degreeUrbanization == 'Suburb: Large'|
                                  degreeUrbanization == 'Suburb: Small'|
                                degreeUrbanization == 'Suburb: Midsize', 'Suburb', degreeUrbanization)
  )
```

```{r}
cleanedDf <- cleanedDf %>%
  mutate(geographicRegion = str_extract(cleanedDf$geographicRegion,  "Far West|Southwest|Rocky Mountains|New England|Mid East|Great Lakes|Plains|Southeast") )
```

Converting  `geographicRegion` and `degreeUrbanization` into factors will be helpful for creating models with a reference group.

```{r}
#converting into factors
cleanedDf <-cleanedDf %>% 
  mutate(geographicRegion_factor = as_factor(geographicRegion), 
         degreeUrbanization_factor = as_factor(degreeUrbanization))

#get rid of US Service Schools
cleanedDf <- cleanedDf%>%
  filter(geographicRegion != 'US Service schools')
```

```{r}
#List all the rows with NA
cleanedDf[rowSums(is.na(cleanedDf)) > 0, ]
```

There are missings for variables that are used in the model. They indicate incompleteness, so dropping them makes sense here.

```{r}
#Deal with NA
library(dplyr)
cleanedDf <- anti_join(cleanedDf, cleanedDf[rowSums(is.na(cleanedDf)) > 0, ])
```

```{r}
write_csv(cleanedDf, "~/Stats 272 F22/Project/Group 7A: Swagat, Zoe, Maya/cleanedDf.csv")
```


